---
title: "Introduction to visualization using ggplot"
author: "Acheampong Andrews"
date: "2023-05-13"
output: html_document
---

Loading the library
```{r}
library(ggplot2)
```

Loading the titanic data set for analysis
```{r}
titan = read.csv("/home/acheampong/Downloads/titanic.csv", stringsAsFactors = FALSE)
```

Setting up factor variables
```{r}
titan$Survived = as.factor(titan$Survived)
titan$Pclass = as.factor(titan$Pclass)
titan$Sex = as.factor(titan$Sex)
titan$Embarked = as.factor(titan$Embarked)
```

1. What was the survival rate?
  * As Survived is the factor (ie. categorical variable), a bar chart is a great visualization to use.
```{r}
ggplot(titan, aes(x = Survived)) +
  theme_bw() +
  geom_bar() +
  labs(y = "Passenger Count",
       title = "Titanic Survival Rate")
```

2. What was the survival rate per gender?
  * We can use color to look at the two aspects (i.e., dimensions) of the data simultaneously
```{r}
ggplot(titan, aes(x = Sex, fill = Survived)) +
  theme_bw() +
  geom_bar() +
  labs(y = "Passenger Count", 
       title = "Titanic Survival Rate by Gender")
```

3. What was the survival rate by class of ticket?
```{r}
ggplot(titan, aes(x = Pclass, fill = Survived)) +
  theme_bw() +
  geom_bar() +
  labs(x = "Passenger Count", 
       title = "Titanic Survival Rate by Ticket Class")
prop.table(table(titan$Survived, titan$Pclass))
```

4. What was the survival rate by class of ticket and gender?
```{r}
ggplot(titan, aes(x = Sex, fill = Survived)) +
  theme_bw() +
  facet_wrap(~ Pclass) +
  geom_bar() +
  labs(y = "Passenger Count",
       title = "Titanic Survival Rate by Pclass and Gender")
prop.table(table(titan$Survived, titan$Pclass, titan$Sex))
```

5. What is the distributions of passenger ages?
  * The histogram is a stable of visualizing numeric data as it very  powerfully communicate the distribution of variable (i.e., columns)
```{r}
ggplot(titan, aes(x = Age)) +
  theme_bw() +
  geom_histogram(binwidth = 5) +
  labs(y = "Passenger Count",
       x = "Age (binwidth = 5)",
       title = "Titanic Age Distribution")
```

6. What are survival rates by age?
```{r}
ggplot(titan, aes(x = Age, fill = Survived)) +
  theme_bw() +
  geom_histogram(binwidth = 5) +
  labs(y = "Passenger Count",
       x = "Age (binwidth = 5)",
       title = "Titanic Survival Rate by Age")
```

Another great visualization for this question is the box-and-wiskey plot
```{r}
ggplot(titan, aes(x = Survived, y = Age)) +
  theme_bw() +
  geom_boxplot() +
  labs( x = "Survived",
        y = "Age",
        title = "Titanic Survival Rate by Age"
  )
```

7. What is the survival rate by age when segmented by gender and class of ticket?
```{r}
ggplot(titan, aes(x = Age, fill = Survived)) +
  theme_bw() +
  facet_wrap(Sex ~ Pclass) +
  geom_density(alpha = 0.5) +
  labs(y = "Age",
       x = "Survived",
       title = "Titanic Survival Rate by Age, Pclass and Sex")
```

We can also use histogram
```{r}
ggplot(titan, aes(x = Age, fill = Survived)) +
  theme_bw() +
  facet_wrap(Sex ~ Pclass) +
  geom_histogram(binwidth = 5) +
  labs(y = "Age",
       x = "Survived",
       title = "Titanic Survival Rate by Age, Pclass and Sex")
```
